---
title: "segmenTree package"
author: "Iyar Lin"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: github_document
---

```{r setup, include=FALSE, purl=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r generate example file, eval=FALSE, echo=F}
knitr::purl("README.Rmd", output = "examples/segmenTree_example.R")
```
The `segmenTree` package implements a decision tree based algorithm for exploration of heterogeneous treatment effects in Randomized Controlled Trial data. It uses the rpart package as the back-end with a predefined method on top.

You can install and load the package from github using:

```{r, purl=FALSE}
pacman::p_load_gh("IyarLin/segmenTree")
```

```{r, echo=FALSE}
library(segmenTree)
```

Let's generate a dataset with a binary response:

```{r generate a dataset}
set.seed(1) # vary seed, n and effect_size below to get a sense of the model performance sensetivity
effect_size <- 0.25
p_x <- function(Tr, X1, X2, X3){
  lp <- 2*X1 + 0.2*X2 + as.numeric(X3)/6
  effect_size + (effect_size/0.125)*Tr*X1^3 + 
    (1 - 2*effect_size)*exp(lp)/(1+exp(lp))
}

n <- 10000
Tr <- rbinom(n, 1, 0.3)
X1 <- runif(n, -0.5, 0.5)
X2 <- rnorm(n)
X3 <- factor(sample(LETTERS[1:3], size = n, replace = T))
p <- p_x(Tr, X1, X2, X3)
y <- sapply(p, function(x) rbinom(1, 1, x))
y_mat <- cbind(y, Tr)
dat <- data.frame(y = I(y_mat), X1, X2, X3)
```

Next, let's import the lift method and use rpart to fit a segment tree:

```{r fit a segment tree}
lift_method <- import_lift_method()
segment_tree <- rpart(y ~ ., data = dat,
                      method = lift_method, 
                      control = rpart.control(cp = 0, minbucket = 1000),
                      x = T)
```

One way of exploring the resulting tree is by printing the rpart object:

```{r explore resulting tree}
segment_tree
```

The `segmenTree` package also contains a utility function to extract the resulting segments:

```{r, warning=F, message=F}
segments <- extract_segments(segment_tree, alpha = 0.15)
print(segments)
```

Next, let's compare how the segment tree predictions (black) compare with the true treatment effect (red):

```{r predict treatment effect and compare with actual treatment effect}
tau <- predict(segment_tree, dat)
p_treat <- p_x(rep(1, n), X1, X2, X3)
p_cont <- p_x(rep(0, n), X1, X2, X3)
cate <- p_treat - p_cont

y_lim <- c(min(tau, cate), max(tau, cate))
plot(c(min(dat$X1), max(dat$X1)), y_lim, type = "n", main = "segmenTree",
     xlab = "X1", ylab = "true (red) vs predicted (black) lift")
points(dat$X1, cate, col = "red")
points(dat$X1, tau)
```

We can see that the algorithm predicts well most of the curve, but did some over-fitting in the middle section.

The `segmenTree::tunecp` function uses cross validation to find the optimal `cp` hyper parameter value.

Below we can see the effective lift and error as a function of the `cp` parameter:

```{r prune tree using tunecp, warning=FALSE}
optimal_cp_cv <- tune_cp(segment_tree)
optimal_cp_cv <- optimal_cp_cv$optimal_cp
pruned_segment_tree <- prune(segment_tree, cp = optimal_cp_cv)
```

The optimal `cp` maximizes the ratio between effective lift and error.

Let's explore the resulting pruned tree segments:

```{r, warning=F, message=F}
segments <- extract_segments(pruned_segment_tree, alpha = 0.15)
print(segments)
```

To demonstrate why we even need a specialized algorithm we'll compare the segmenTree model with 2 other approaches that use plain ML models:

1. Model jointly the treatment and covariates. Predict: tau(x)=f(x,T=1)-f(x,T=0)  
1. Train a model on the treatment units f_{T=1} and a separate model f_{T=0} on the control units and predict: tau(x) = f_{T=1}(x) - f_{T=0}(x)  

We'll use a decision tree in both of the above.

```{r compare segmenTree with 2 other approches}
# segmenTree pruned model
tau <- predict(pruned_segment_tree, dat)
p_treat <- p_x(rep(1, n), X1, X2, X3)
p_cont <- p_x(rep(0, n), X1, X2, X3)
cate <- p_treat - p_cont

y_lim <- c(min(tau, cate), max(tau, cate))
plot(c(min(dat$X1), max(dat$X1)), y_lim, type = "n", main = "segmenTree",
     xlab = "X1", ylab = "true (red) vs predicted (black) lift")
points(dat$X1, cate, col = "red")
points(dat$X1, tau)

# Model jointly the treatment and covariates
dat2 <- data.frame(y, X1, X2, X3, Tr)
fit2 <- rpart(y ~ ., data = dat2)

dat2_treat <- dat2; dat2_cont <- dat2
dat2_treat$Tr <- 1L; dat2_cont$Tr <- 0L
tau2 <- predict(fit2, dat2_treat) - predict(fit2, dat2_cont)

y_lim <- c(min(tau2, cate), max(tau2, cate))
plot(c(min(dat$X1), max(dat$X1)), y_lim, type = "n", main = "regular model",
     xlab = "X1", ylab = "")
points(dat$X1, cate, col = "red")
points(dat$X1, tau2)

# Train a model on the treatment units and a seperate model on the control units

dat3_treat <- dat2[dat2$Tr == 1, -5]
dat3_cont <- dat2[dat2$Tr == 0, -5]

fit3_treat <- rpart(y ~ ., data = dat3_treat)
fit3_cont <- rpart(y ~ ., data = dat3_cont)
dat2_treat <- dat2; dat2_cont <- dat2
tau3 <- predict(fit3_treat, dat2) - predict(fit3_cont, dat2)

y_lim <- c(min(tau3, cate), max(tau3, cate))
plot(c(min(dat$X1), max(dat$X1)), y_lim, type = "n", main = "separate models",
     xlab = "X1", ylab = "")
points(dat$X1, cate, col = "red")
points(dat$X1, tau3)
```
